hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-03-07 inst/manuscript/paper.Rmd
final copy edits and render

In this scenario, we consider road work near the upper portion of Crooked River in the Boise River watershed, home to a key Bull Trout (*Salvelinus confluentus*) population. Bull Trout, a federally threatened species under the Endangered Species Act [@usfws_bt_1999], are particularly sensitive to sediment disturbance [@nolfi2024]. The Forest Service Fisheries Biologist overseeing the project wants to determine whether Bull Trout fry will likely be out of the gravel and free-swimming by June 1st. In this system, Bull Trout typically complete spawning by the end of September, so we consider the latest possible spawn date: September 30th.

@@

# predict emergence time using Sept. 30 as the spawn date
bt_emergence <- predict_phenology(
  data = EFS_data,
  dates = date,
  temperature = temp_c,
  spawn.date = "2011-09-30",
  model = bt_emerge_mod
)

@@

In this example, the model predicts that the last Bull Trout will emerge on May 1, before the road work target date (June 1). This suggests that the Fisheries Biologist can confidently approve the work in the area without concern for sediment disturbance impacting fish developing in the gravel. This type of predictive modeling helps managers make informed, science-based decisions that balance conservation priorities with land-use activities.

@@

Second, developmental timing occurs as a distribution, not a fixed point. While **hatchR** provides point estimates for developmental phenology, fish spawning and development within populations occur as distributions rather than single events [@mason1976]. We encourage users to predict phenology using early, peak, and late thresholds (e.g., 5th, 50th, and 95th percentiles) or incorporate real or modeled distributions to capture variation.

@@

Importantly, hatchR extends the effective value framework developed by @sparks2019 into a generalizable tool that can be applied to any fish species or population, provided that appropriate source data are available—data that can easily be collected in aquaculture settings. We present foundational applications of **hatchR**, with additional user-friendly cases and implementation guides available on the software's website. The software is designed for both applied and fundamental research, allowing users to engage with it either through a programmatic R environment or via a user-friendly Shiny app. We expect that the examples provided here represent only a fraction of **hatchR**'s potential applications and encourage the user community to explore and expand upon this framework for their own research and management needs.

>>>>

In this scenario, we consider road work near the upper portion of Crooked River in the Boise River watershed, home to a key Bull Trout (*Salvelinus confluentus*) population. Bull Trout, a federally threatened species under the Endangered Species Act [@usfws_bt_1999, @nolfi2024], are particularly sensitive to sediment disturbance. The Forest Service Fisheries Biologist overseeing the project wants to determine whether Bull Trout fry will likely be out of the gravel and free-swimming by June 1st. In this system, Bull Trout typically complete spawning by the end of September, so we consider the latest possible spawn date: September 30th.

@@

# predict emergence time using Sept. 30 as the spawn date
bt_emergence <- predict_phenology(
  data = EFS_data,
  dates = date,
  temperature = temp_c,
  spawn.date = "2014-09-30",
  model = bt_emerge_mod
)

@@

In this example, the model predicts that the last Bull Trout will emerge on April 21, before the road work target date (June 1). This suggests that the Fisheries Biologist can confidently approve the work in the area without concern for sediment disturbance impacting fish developing in the gravel. This type of predictive modeling helps managers make informed, science-based decisions that balance conservation priorities with land-use activities.

@@

Second, developmental timing occurs as a distribution, not a fixed point. While **hatchR** provides point estimates for developmental phenology, fish spawning and development within populations occur as distributions rather than single events [@mason1976]. We encourage users to predict phenology using early, peak, and late thresholds (*e.g.*, 5th, 50th, and 95th percentiles) or incorporate real or modeled distributions to capture variation.

@@

Importantly, hatchR extends the effective value framework developed by @sparks2019 into a generalizable tool that can be applied to any fish species or population, provided that appropriate source data are available. We present foundational applications of **hatchR**, with additional user-friendly cases and implementation guides available on the software's website. The software is designed for both applied and fundamental research, allowing users to engage with it either through a programmatic R environment or via a user-friendly Shiny app. We expect that the examples provided here represent only a fraction of **hatchR**'s potential applications and encourage the user community to explore and expand upon this framework for their own research and management needs.
